Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China

### 2024-02-02 analysis/paper/paper.qmd
work on result

@@

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China. As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting. The post-excavation analysis of Guijiabao is still in progress. This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology. We aim to explore the differences in overall microbial communities, including bacterial, archaeal and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples. Our main research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? We hypothesized that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective could shed light on food practices from a microbial perspective that has not been fully discussed in archaeology.

@@

To minimize potential sources of contamination from the environments, we cleaned all lab benches, equipment, and pipettors with 3% bleach, followed by 99% ethanol and UV light. The sample preparation and DNA extraction were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan. Because modern contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions, reagents, and consumables. A total of 12 powdered pottery samples (interior and exterior parts), 5 soil samples from Guijiabao and Gaoshan, and 3 negative control samples were analyzed in this study.

@@

Given the considerable fragmentation of DNA in archaeological contexts, we focused on short regions ranging from 100 to 300 bp, including V1, V4, and V6 of 16S rRNA and ITS1 of 18S rRNA. For bacterial communities, the V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@]. For archaeal communities, V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Siles2018]. For fungal communities, ITS1 region of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018]. Barcoded forward adapters were added to the PCR primers to facilitate sample pooling for library preparations. The DNA amplification for bacteria was conducted using the Taq DNA polymerase Master Mix RED (Ampliqon) under the following program: initial denaturation at 95 °C for 1 min, followed by 35 cycles of denaturation at 95 °C for 30 s, primer annealing at 50 °C for 30 s, extension at 72 °C for 30 s, and final elongation at 72 °C for 5 min. For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and annealing temperature at 57 °C for 40 s.

@@

@@

@@

```{r}
#| label: get-data
#| eval: false
# Note the path that we need to use to access our data files when rendering this document
my_data <- read.csv(here::here('analysis/data/raw_data/my_csv_file.csv'))
```

>>>>

```{r setting}
#| echo: false
# Format integers
options(scipen=999)
knit_print.integer = function(x, ...) {
  prettyNum(x, big.mark=",")
}

registerS3method(
  "knit_print", "integer", knit_print.integer,
  envir = asNamespace("knitr")
)

```

@@

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China. As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting. The post-excavation analysis of Guijiabao is still in progress. This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology. We aim to explore the differences in overall microbial communities, including bacterial and fungal communities, through a comparative analysis of different parts of ceramic samples and soil samples. Our main research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? We hypothesized that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective could shed light on food practices from a microbial perspective that has not been fully discussed in archaeology.

@@

To minimize potential sources of contamination from the environments, we cleaned all lab benches, equipment, and pipettors with 3% bleach, followed by 99% ethanol and UV light. The sample preparation and DNA extraction were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan. Because modern contamination could be a significant concern when interpreting the data, we used negative controls during sample preparation, and reagent blanks for DNA extraction, and PCR amplification to monitor extraneous DNA from surroundings, reagents, and consumables. A total of 12 powdered pottery samples (interior and exterior parts), 5 soil samples from Guijiabao and Gaoshan, and 3 negative control samples were analyzed in this study.

@@

Given the fragmentation of DNA in archaeological contexts, we targeted single hypervariable regions, V1, V4, and V6 of 16S rRNA, and ITS1 of 18S rRNA, to cover both bacterial and fungal communities. The V1 region was amplified using the primers 8F (5'-AGAGTTTGATYMTGGCTCAG-3') and 120R (5'-TTACTCACCCGTICGCCRCT-3') [@Adler2013], the V4 region was amplified using arch519F (5′-CAGCMGCCGCGGTAA-3′) and arch806R (5′-GGACTACVSGGGTATCTAAT-3′), and the V6 region was amplified using 926F (5’-AAACTYAAAKGAATTGACGGG-3’) and 1046R (5’-CGACARCCATGCASCACCT-3’) [@Adler2013; @Siles2018]. For fungal communities, the region ITS1 of 18S rRNA was amplified using the primers ITS1F (5’-CTTGGTCATTTAGAGGAAGTAA-3’) and ITS2 (5′-GCTGCGTTCTTCATCGATGC-3′) [@Siles2018]. Barcoded forward adapters were added to the PCR primers to facilitate sample pooling for library preparations. The DNA amplification for bacteria was conducted using the Taq DNA polymerase Master Mix RED (Ampliqon) under the following program: initial denaturation at 95 °C for 1 min, followed by 35 cycles of denaturation at 95 °C for 30 s, primer annealing at 50 °C for 30 s, extension at 72 °C for 30 s, and final elongation at 72 °C for 5 min. For fungi, the thermal conditions are the same as for bacteria, with the exception of 40 cycles and primer annealing temperature at 57 °C for 40 s.

@@

## DNA library construction and sequencting

## Taxonomic classification

@@

```{r get-data}
#| label: get-data
#| echo: false
#| cache: true
library(tidyverse)
library(readxl)
# read original count data
OTU_V4 <- read.table(here::here("analysis","data","raw_data","V4_asv_table.txt"), header = T, sep="\t")
OTU_V1 <- read.table(here::here("analysis","data","raw_data","V1_asv_table.txt"), header = T, sep="\t")
OTU_V6 <- read.table(here::here("analysis","data","raw_data","V6_asv_table.txt"), header = T, sep="\t")
OTU_ITS1 <- read.table(here::here("analysis","data","raw_data","ITS1_asv_table.txt"), header = T, sep="\t")

OTU_reads_total <- function(df) {
  sum_read <- df %>%
    rowwise() %>% 
    mutate(sum_read = sum(across(where(is.numeric)))) %>% 
    select(sum_read)
  return(sum(sum_read))
}

V4_total_reads <- OTU_reads_total(OTU_V4)
V1_total_reads <- OTU_reads_total(OTU_V1)
V6_total_reads <- OTU_reads_total(OTU_V6)
ITS1_total_reads <- OTU_reads_total(OTU_ITS1)
total_reads <- V4_total_reads + V1_total_reads + V6_total_reads + ITS1_total_reads

# read rarefied data
rarefy_V4 <- read.csv(here::here("analysis","data","derived_data", "V4_shared_format.1.subsample.60216.csv"), sep = ",")
rarefy_V1 <- read.csv(here::here("analysis","data","derived_data","V1_shared_format.1.subsample.13658.csv"), sep = "")
rarefy_V6 <- read.csv(here::here("analysis","data","derived_data", "V6_shared_format.1.subsample.28402.csv"), sep = ",")
rarefy_ITS1 <- read.csv(here::here("analysis","data","derived_data", "ITS1_shared_format.1.subsample.387.csv"), sep = ",")

rarefy_reads <- function(df) {
  sum_read <- df %>%
    rowwise() %>% 
    mutate(sum_read = sum(across(starts_with("ASV")))) %>% 
    select(sum_read)
  return(sum(sum_read))
}

V4_rare_reads <- rarefy_reads(rarefy_V4)
V1_rare_reads <- rarefy_reads(rarefy_V1)
V6_rare_reads <- rarefy_reads(rarefy_V6)
ITS1_rare_reads <- rarefy_reads(rarefy_ITS1)
rare_reads <- V4_rare_reads + V1_rare_reads + V6_rare_reads + ITS1_rare_reads

V4_OTU_num <- unique(rarefy_V4$numOtus)
V1_OTU_num <- unique(rarefy_V1$numOtus)
V6_OTU_num <- unique(rarefy_V6$numOtus)
ITS1_OTU_num <- unique(rarefy_ITS1$numOtus)
total_OTU <- V4_OTU_num + V1_OTU_num + V6_OTU_num + ITS1_OTU_num
```

## Alpha diversity analysis of bacterial and fungal communities
Illumina analysis generated a total of `r total_reads` sequences that were normalized to `r rare_reads` after rarefaction [@Schloss2024]. Soil samples have the highest shannon diversity index and richness, followed by pottery exterior samples, and pottery interior samples, and negative control.

## Taxonomic composition

## Beta diversity analysis

@@
